---
title: "Exploring lakefish data"
author: "Emma Sofie Skarstein and Lyder Iversen"
date: "November 2019"
output: bookdown::html_document2
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height = 8, 
                      echo=FALSE, warning=FALSE, message=FALSE)
```

```{r libraries}
library(ggplot2)
library(sf) # for sf objects
library(dplyr) # for smoother dataframe-manipulation
library(gganimate)
library(maps)
library(maptools)
library(mapdata)
library(viridis)
library(cowplot)
```

## About the data

We will be looking at observations of freshwater fish in Norwegian lakes. Our observations are obtained from GBIF (more formal referencing of data somehow?), by first downloading all ray-finned fish (Actinopterygii) observations registered in Norway, and then filtering these based on the following steps:

1. Filter out species not in list of Norwegian freshwater fish, [as described by SNL](https://snl.no/ferskvannsfisk_i_Norge) 
2. Match all species to closest lake, and remove observation further than 10 meters from the closest lake
3. Remove all observations with no time variable.
4. Select 12 most prevalent species. 


The list of fish we downloaded can be seen below. We excluded fish that spawn in salt water, so some species that may be observed in lakes, such as eel (_Anguilla anguilla_) and European flounder (_Platichthys flesus_), have been excluded from our analysis as a result of this. 

```{r fishnames}
fish_names <- read.table("fish_names.txt", sep = "\n", stringsAsFactors = FALSE)[,1]
cat(fish_names, sep = ", ")
```

```{r loading_data}
# Rmarkdown ser ut til å ha problemer med å lokalisere datafilene som ligger i en annen mappe enn selve mappa der rapporten ligger, så jeg legger inn de data-filene vi trenger der

# Loading filtered observations
occ <- readRDS("filtered_occurences.rds")
#lakes <- readRDS("lake_polygons.rds")
```

## Preliminary insights

To begin with, for the sake of us non-biologists, here are the latin, English and Norwegian names of the twelve included species.

* Phoxinus phoxinus - Minnow - Ørekyte
* Rutilus rutilus - Roach - Mort      
* Salmo trutta - Trout - Ørret
* Esox lucius - Pike - Gjedde        
* Salmo salar - Salmon - Laks          
* Gasterosteus aculeatus - Three-spined stickleback - Trepigget stingsild
* Lota lota - Burbot - Lake
* Perca fluviatilis - Perch - Abbor
* Salvelinus alpinus - Arctic char - Røye  
* Coregonus lavaretus - Lavaret - Sik
* Coregonus albula - European cisco - Lagesild
* Salvelinus fontinalis - Brook char - Bekkerøye

Let us take a look at the number of observations of each species. 

```{r species_barplot, echo = FALSE}
# Bar plot of different species
species_counts <- count(occ, vars = species)
ggplot(species_counts, aes(x = vars, y = n, fill = vars)) + 
  geom_bar(stat = "identity", color="black") + 
  geom_text(aes(label = n), vjust=-0.3, color="black", size=4) +
  theme_light() + 
  ylab("Number of observations") +
  theme(text = element_text(size = 15), 
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank()) +
  scale_fill_brewer(palette="Paired") 
  
```

Here we see that _Salmo trutta_ (trout) outranks the other species by quite a bit, although _Perca fluviatilis_ (European perch) and _Salvelinus alpinus_ (arctic char) are also quite prevalent. 

It is also interesting to take a look at when the observations were made. An important characteristic of our data is the fact that the observations appear at very irregular times, there are some years when there seem to be large studies of many lakes, while other years there are only a few single observations. The earliest observation is made in `r min(occ$year)`, but the first decades are mostly just observations made in single lakes and are not very numerous. We look at the observation counts for the years after 1970, which is when observations really start increasing.

```{r years_barplot}
recent_occ <- occ %>% filter(year >= 1970)
time_counts_recent <- count(recent_occ, year, species)

ggplot(time_counts_recent, aes(x = year, y = n, fill = species)) + 
  geom_bar(stat = "identity", color="black", position = position_stack(reverse = FALSE)) + 
  theme_light() + 
  ylab("Number of observations") + 
  scale_fill_brewer(palette="Paired")
  
```

So 1993 was the big lakefish year, especially for _Salmo trutta_ and _Salvelinus alpinus_. It is also interesting to note that there are relatively many observations of _Coregunus albula_ and _Coregonus lavaretus_ during the period 2002 to 2005, and barely any at all the other years. The _Perca fluviatilis_ is also interesting in the sense that the species are fairly stable over time, there are quite a few observations every year, but never any huge spikes such as _S. trutta_ and _S. alpinus_ have in 1993.

Lastly, we also take a look at where the observations are made.
```{r fylke_barplot}
dense_fylke <- count(occ, vars = county.y)
colnames(dense_fylke) <- c("county.y", "obs_in_county", "geometry")
ggplot(dense_fylke, aes(x = county.y, y = obs_in_county)) + 
  geom_bar(stat = "identity", color="black", fill="antiquewhite") + 
  geom_text(aes(label = obs_in_county), vjust=-0.3, color="black", size=4) +
  theme_light() + 
  ylab("Number of observations") +
  theme(text = element_text(size = 15), 
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank())
```

## Plotting data on the map
First, let us look at all the observations on a map of Norway.
```{r all_obs_map}
norway <- ggplot2::map_data("world", region = "Norway(?!:Svalbard)") 
# Set theme for blue ocean
theme_set(theme_light() + theme(aspect.ratio = .70, panel.background = element_rect(fill = "aliceblue")))

ggplot(occ) +
  geom_map(data = norway, map = norway, aes(long, lat, map_id=region), 
           color="#2b2b2b", fill = "white") + 
  geom_point(aes(x = decimalLongitude, y = decimalLatitude, color = species), alpha = 0.6, size = 0.5) + 
  scale_color_brewer(palette="Paired") +
  theme(axis.title.x = element_blank(), 
        axis.title.y = element_blank()) +
  guides(colour = guide_legend(override.aes = list(size=2)))

```

So, we can definitely see _something_ here, but it is a bit of a mess. We have added some transparency to the points, but they are still very much on top of each other, so this does not necessarily convey what we want it to. Let us facet it into one map per species to tidy it up a bit.

```{r all_obs_faceted, fig.width = 12, fig.height = 12}
ggplot(occ) +
  geom_map(data = norway, map = norway, aes(long, lat, map_id=region), 
           color="#2b2b2b", fill = "white") + 
  geom_point(aes(x = decimalLongitude, y = decimalLatitude), alpha = 0.6, size = 0.1, color = "dodgerblue3") + 
  theme(axis.title.x = element_blank(), 
        axis.title.y = element_blank()) +
  guides(colour = guide_legend(override.aes = list(size=2))) + 
  facet_wrap(~ species, nrow = 4)
```

(Comments on this plot)

In order to introduce another dimension here, let us look at the data animated over time.

```{r all_obs_animated}
a <- ggplot(occ) +
  geom_map(data = norway, map = norway, aes(long, lat, map_id=region), 
           color="#2b2b2b", fill = "white") + 
  geom_point(aes(x = decimalLongitude, y = decimalLatitude, color = species), alpha = 0.6, size = 1) + 
  scale_color_brewer(palette="Paired") +
  theme(#legend.text = element_text(size=7), 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank()) +
  guides(colour = guide_legend(override.aes = list(size=2))) + 
  labs(title = paste0("All freshwater fish", ', year: {frame_time}')) + 
  transition_time(year)

animate(a, duration = length(unique(occ$year))*0.5, end_pause = 4)

```

(comments on animation of all species)

Again, let us facet this by species in order to see a bit more detail.

```{r all_obs_animated_faceted, fig.width = 12, fig.height = 12}
a <- ggplot(occ) +
  geom_map(data = norway, map = norway, aes(long, lat, map_id=region), 
           color="#2b2b2b", fill = "white") + 
  geom_point(aes(x = decimalLongitude, y = decimalLatitude), alpha = 0.6, size = 1, color = "dodgerblue3") + 
  theme(axis.title.x = element_blank(), 
        axis.title.y = element_blank()) +
  guides(colour = guide_legend(override.aes = list(size=2))) +
  facet_wrap(~ species, nrow = 4) + 
  labs(title = paste0("All freshwater fish", ', year: {frame_time}')) + 
  transition_time(year)

animate(a, duration = length(unique(occ$year))*0.5, end_pause = 4)
```

(comment on animation faceted by species)
It is slightly worrying to note here that there seem to be observations in Sweden. This is especially worrying since this was not the case in the non-faceted animation. This should be examined further.

When we looked at the bar plot of observations per year at the beginning of this report, we noted that _Perca fluviatilis_ was one of few species that had fairly stable observation numbers across years. It may be interesting to view these observations in an animation.

```{r perca_fluviatilis}
# Code for _Perca fluviatilis_ animation
a <- ggplot(occ %>% filter(species == "Perca fluviatilis")) +
  geom_map(data = norway, map = norway, aes(long, lat, map_id=region), 
           color="#2b2b2b", fill = "white") + 
  geom_point(aes(x = decimalLongitude, y = decimalLatitude), alpha = 0.6, size = 1, color = "dodgerblue3") + 
  theme(axis.title.x = element_blank(), 
        axis.title.y = element_blank()) +
  guides(colour = guide_legend(override.aes = list(size=2))) +
  labs(title = paste0("Perca fluviatilis (perch)", ', year: {frame_time}')) + 
  transition_time(year)

animate(a, duration = length(unique(occ$year))*0.5, end_pause = 4)
```
```{r}
walking_fluviatilis <- occ %>% filter(species == "Perca fluviatilis") %>% filter(year)
```

In the animations above we got a sense of the observations over time, but the constant movement of the animation makes it a bit difficult to absorb. Here we have divided all the observations from 1970 to 2010 into four sections: 1971 - 1980, 1981 - 1990, 1991 - 2000 and 2001 - 2010, and we plot all species for each time period.

```{r all_obs_four_periods}
pick <- function(condition){
  function(d) d %>% filter(!!enquo(condition))
}

norway1980 <- ggplot(data = occ, aes(x = decimalLongitude, y = decimalLatitude)) +
  geom_map(data = norway, map = norway, aes(long, lat, map_id=region), color="#2b2b2b", fill="white") +
  geom_point(data = pick(year<=1980 & year>1970), size=0.001, color = "red3") +
  ggtitle("Fish data between 1970 and 1980")

norway1990 <- ggplot(data = occ, aes(x = decimalLongitude, y = decimalLatitude)) +
  geom_map(data = norway, map = norway, aes(long, lat, map_id=region), color="#2b2b2b", fill="white") +
  geom_point(data = pick(year<=1990 & year>1980), size=0.001, color = "red3") +
  ggtitle("Fish data between 1980 and 1990")

norway2000 <- ggplot(data = occ, aes(x = decimalLongitude, y = decimalLatitude)) +
  geom_map(data = norway, map = norway, aes(long, lat, map_id=region), color="#2b2b2b", fill="white") +
  geom_point(data = pick(year<=2000 & year>1990), size=0.001, color = "red3") +
  ggtitle("Fish data between 1990 and 2000")

norway2010 <- ggplot(data = occ, aes(x = decimalLongitude, y = decimalLatitude)) +
  geom_map(data = norway, map = norway, aes(long, lat, map_id=region), color="#2b2b2b", fill="white") +
  geom_point(data = pick(year<=2010 & year>2000), size=0.001, color = "red3") +
  ggtitle("Fish data between 2000 and 2010")

plot_grid(norway1980, norway1990, norway2000, norway2010, labels = "AUTO")
```

From these plots we see that a large portion of the observations were done between 1990 and 2000, which is expected after seeing the earlier histogram with the large spike in 1993. There also seems to be a pattern with a large area in the middle of southern Norway having close to no observations in this time period. This area does however seem to have the majority of the observations in the next time period. This could indicate that some type of organized survey or observation effort was done in this time period, perhaps to fill this geographic hole in the data.

By counting the number of unique species per lake we can find the species richness per lake. 
```{r species_richness_prep}
occ_w_lake <- occ[complete.cases(occ$waterBodyID),]
rich_lakes <- occ_w_lake %>% group_by(waterBodyID) %>% summarize(species_richness = n_distinct(species))
colnames(rich_lakes) <- c("waterBodyID", "species_richness", "geometry")
rich_lakes <- st_set_geometry(rich_lakes, NULL)
rich_lakes_w_loc <- merge(rich_lakes, occ, by = "waterBodyID")
```
```{r species_richness}
ggplot(rich_lakes_w_loc) + 
  geom_map(data = norway, map = norway, aes(long, lat, map_id=region), 
           color="#2b2b2b", fill = "white") + 
  geom_point(aes(x = decimalLongitude, y = decimalLatitude, color = species_richness), alpha = 0.5, size = 1) +
  scale_colour_viridis(option = "magma", direction = -1) +
  ggtitle("Species richness per lake")
```


(comment on this, maybe with reference to barplot of fylke)

```{r, eval = TRUE}
rich_list <- unique(data.frame(waterBody = rich_lakes_w_loc$waterBody, species_richness = rich_lakes_w_loc$species_richness))
arrange(rich_list, by_group = desc(species_richness))[1:10,]
```
